---
title: "data_cleaning"
author: "Trenten"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

**Data Wrangling**: The following tasks with be completed:

- Clean data collected from webscraping VASE website. 
- Left join VASE dataset with TEA's dataset. 
- Lastly, join new data setsets with geospatial data.

VASE and TEA data will be simultaneously worked on. Geospatial data will be incorpated at the very end.

```{r message = FALSE}
vase <- fread("VASE_rawdata.csv")
tea22 <- fread("TEA22_rawdata.csv")
```

**Data Setup**: Datasets will be manipulated to only contain values of interest. Obviously unuseable names (that are not a school or too vague for identification purposes) that cannot be linked to a high school will be discarded.

```{r}
# check variable names
colnames(vase)
colnames(tea22)

# check for extremely unusual values
#unique(vase$School)

# rename column names
vase <- vase %>%
  select(-c("Student", "Title")) %>%
  rename("Art_Region" = Region) %>%
  mutate("Region" = NA) %>%
  filter(!School %in% c("Gandrews@aledo.k12.tx.us", "Home School",
                        "Homeschool/seton Home Study", "High School",
                        "High School, Junior H S"))

tea22 <- tea22 %>%
  select(CAMPNAME, CPE0312C,
         GRDHIGH, REGION, DISTNAME) %>%
  rename("School" = CAMPNAME,
         "Enrolled" = CPE0312C,
         "Highest_Grade" = GRDHIGH,
         "Region" = REGION,
         "District" = DISTNAME) %>%
  filter(Highest_Grade %in% c(9, 10, 11, 12)) #reduce size of df and repetitiveness of school names

tea22 <- tea22 %>%
  select(-Highest_Grade)
```

**Preliminary Data Standardization**: School names in `vase` will be somewhat standardized. Region information will also be standardized. This preliminary standardizations will assist with later fuzzy joining.

```{r}
# standardize Region distinctions
# remove subgrouping of region to match with TEA
for (i in (1:length(vase$Art_Region))){
  if (substring(vase$Art_Region[i], nchar(vase$Art_Region[i]),nchar(vase$Art_Region[i])) %in% c("N", "E", "S", "W")){
    vase$Region[i] <- substring(vase$Art_Region[i],1,nchar(vase$Art_Region[i])-1)
  }
  else{
    vase$Region[i] <- vase$Art_Region[i]
  }
}

# remove leading zeros and apostrophe of Region to match with VASE
for (i in (1:length(tea22$Region))){
  tea22$Region[i] <- gsub("^\\'0|^\\'", "", tea22$Region[i])
}

# standardize VASE HS names in similar format as TEA22
# clean HS and ECHS at end of names first

# function to clean up endings
re_string <- function(vect, suffixes, ending){
  column <- vect
  for (i in suffixes){
    for (j in 1:length(column)){
      word <- paste("",i)
      if(word == substring(column[j],nchar(column[j])-nchar(word)+1,nchar(column[j]))){
        a = substring(column[j],1,nchar(column[j])-nchar(word))
        column[j] <- paste0(a, ending)
      }
    }
  }
  return(column)
}

un_string <- function(vect, suffixes){
  column <- vect
  for (i in suffixes){
    for (j in 1:length(column)){
      word <- i
      if(word == substring(column[j],nchar(column[j])-nchar(word)+1,nchar(column[j]))){
        a = substring(column[j],1,nchar(column[j])-nchar(word))
        column[j] <- str_trim(a, "right")
      }
    }
  }
  return(column)
}

# endings to clean up
end_hs <- c("HS", "hs", "High Schoo.l", 
          "High School", "Hs", "Highschool",
          "H.s.", "High", "High-school",
          "H. S.", "H", "High Schooll")

end_echs <- c("E C H S", "Echs", "Early College HS",
             "Early College High School", "Early College H S",
             "EARLY COLLEGE H S", "EARLY COLLEGE HS", "Early College",
             "EARLY COLLEGE HIGH SCHOOL")

end_extra <- c("Faa", ", Reg. 16", ", Reg. 20w",
               "9", "Home", "ISD", "Isd", "9th", "9th Grade Campus", 
               "Ninth Grade Center", 
               "9th Grade Center", "Harmon Campus", "Harmon")

vase$School <- un_string(vase$School, end_extra)
vase$School <- re_string(vase$School, end_hs, " H S")
vase$School <- re_string(vase$School, end_echs, " ECHS")
tea22$School <- re_string(tea22$School, end_echs, " ECHS")

# check results of preliminary standardization
#sort(unique(vase$School))

# edit manually the following values to improve fuzzy join
# school names are not unique, sometimes even within regions
# Seemingly no convention between un- and official names

# function to rename certain schools
re_name <- function(orig, neww, reg){
  temp <- vase
  temp$School[temp$School == orig & temp$Region == reg] <- neww
  return(temp)
}

vase <- re_name("Aldine Eisenhower", "EISENHOWER H S", 4)
vase <- re_name("Aldine Mac Arthur", "MACARTHUR H S", 4)
vase <- re_name("Aldine Macarthur H S", "MACARTHUR H S", 4)
vase <- re_name("Aldine Carver", "CARVER H S FOR APPLIED TECH/ENGINEERING/ARTS", 4)
vase <- re_name("Harmony Science Academy North Austin", "HARMONY SCIENCE ACADEMY-PFLUGERVILLE", 13)
vase <- re_name("Mozart Secondary", "VANGUARD MOZART", 1)
vase <- re_name("L B J H S", "LBJ ECHS", 13)
vase <- re_name("L B J H S", "LBJ ECHS", 13)
vase <- re_name("L B J Johnson City", "LBJ ECHS", 13)
vase <- re_name("I. M. Terrell H S", "IM TERRELL FOR STEM AND VPA", 11)
vase <- re_name("L.d. Bell H S", "BELL H S", 11)
vase <- re_name("Wunsche", "CARL WUNCSCHE SR H S", 4)
vase <- re_name("Carver", "CARVER H S FOR APPLIED TECH/ENGINEERING/ARTS", 4)
vase <- re_name("Barack Obama Male Leadership Academy","BARACK OBAMA MALE LEADERSHIP ACAD AT A MOCEO SMITH",10)
vase <- re_name("Booker T. Washington H S", "BOOKER T WASHINGTON SPVA MAGNET", 10)
vase <- re_name("Booker T Washington H S", "BOOKER T WASHINGTON SPVA MAGNET", 10)
vase <- re_name("Benbrook Middle-hs", "BENBROOK MIDDLE/HIGH SCHOOL", 11)
vase <- re_name("Benbrook Middle H S", "BENBROOK MIDDLE/HIGH SCHOOL", 11)
vase <- re_name("Everman Joe C Bean H S","EVERMAN H S", 11)
vase <- re_name("Santa Anna", "SANTA ANNA SECONDARY", 15)
vase <- re_name("James Pace H S", "PACE ECHS", 1)
vase <- re_name("Roy Miller H S", "ROY MILLER H S AND METRO SCHOOL OF DESIGN", 2)
vase <- re_name("Memorial ECHS", "VETERANS MEMORIAL ECHS", 1)
vase <- re_name("Memorial ECHS", "EASTSIDE MEMORIAL ECHS", 13)
vase <- re_name("Avalos Tech", "AVALOS P-TECH SCHOOL", 4)
vase <- re_name("Blanson", "BLANSON CTE H S", 4)
vase <- re_name("Connally H S", "JOHN B CONNALLY H S", 13)
vase <- re_name("Connally H S", "CONNALLY HIGH SCHOOL", 12)
vase <- re_name("Cumberland Academy H S", "CUMBERLAND H S", 7)
vase <- re_name("Cumberland Academy", "CUMBERLAND H S", 7)
vase <- re_name("Lake Ridge H S", "MANSFIELD LAKE RIDGE H S", 11)
vase <- re_name("Timberview H S", "MANSFIELD TIMBERVIEW H S", 11)
vase <- re_name("Richard King H S", "KING H S", 2)
vase <- re_name("Champion H S", "BOERNE - SAMUEL V CHAMPION H S", 20)
vase <- re_name("Marshall H S", "THURGOOD MARSHALL H S", 4)
vase <- re_name("Summit H S", "Mansfield Summit H S", 11)
vase <- re_name("Bryan H S", "TRAVIS B BRYAN H S", 6)
vase <- re_name("Elkins H S", "LAWRENCE E ELKINS H S", 4)
vase <- re_name("Heath H S", "ROCKWALL-HEATH H S", 10)
vase <- re_name("Legacy H S", "Mansfield Legacy H S", 11)
vase <- re_name("Legacy H S", "TYLER LEGACY H S", 7)
vase <- re_name("Travis H S", "WILLIAM B TRAVIS H S", 4)
vase <- re_name("Travis H S", "LAKE TRAVIS H S", 13)
vase <- re_name("Dawson H S", "GLENDA DAWSON H S", 4)
vase <- re_name("Hawkins H S", "HAWKINS H S / MIDDLE", 7)
vase <- re_name("Lee H S", "LEGACY TRADITIONAL SCHOOL - LEE ACADEMY", 20)
vase <- re_name("Molina H S", "MOISES E MOLINA H S", 10)
vase <- re_name("Wagner H S", "KAREN WAGNER H S", 20)
vase <- re_name("Braswell H S", "RAY BRASWELL H S", 11)
vase <- re_name("Eaton H S", "V R EATON H S", 11)
vase <- re_name("Terry H S", "B F TERRY H S", 4)
vase <- re_name("Lassiter ECHS", "DR WRIGHT L LASSITER JR ECHS", 10)
vase <- re_name("Baytown Lee", "LEE H S", 4)
vase <- re_name("Vehs", "VICTORIA EAST H S", 3)
vase <- re_name("Alameda School For Art + Design", "HENRY FORD ACADEMY ALAMEDA SCHOOL FOR ART + DESIGN", 20)
vase <- re_name("Pegasus School Of Liberal Arts & Sciences", "PEGASUS CHARTER H S", 10)
vase <- re_name("Texas High School-texarkana", "	
TEXAS H S", 8)
vase <- re_name("Lewisville High School Killough", "	
LEWISVILLE H S", 11)
vase <- re_name("Harlan High School / Nisd", "HARLAN H S", 20)
vase <- re_name("Uplift North Hills Preparatory", "UPLIFT EDUCATION-NORTH HILLS PREP H S", 10)
vase <- re_name("Lhs Killough", "LEWISVILLE H S", 11)
vase <- re_name("The Woodlands College Park", "THE WOODLANDS H S", 6)
vase <- re_name("The Woodlands College Park H S", "THE WOODLANDS H S", 6)
vase <- re_name("Meridian School", "MERIDIAN WORLD SCHOOL LLC", 13)
vase <- re_name("Palmview H S", "LA JOYA PALMVIEW H S", 1)
vase <- re_name("Shoemaker H S","ROBERT M SHOEMAKER H S",12)
vase <- re_name("Riesel", "RIESEL SCHOOL", 12)
vase <- re_name("Rudder H S", "JAMES EARL RUDDER H S", 6)
vase <- re_name("Centennial H S", "BURLESON CENTENNIAL H S", 11)
vase <- re_name("Davis", "DAVIS H S ALDINE", 4)
vase <- re_name("Leakey H S", "LEAKEY SCHOOL", 20)


# Schools that appeared in VASE, but do not actually exist 
# within stated region or are not considered public under 
# TEA (but they had close matches in fuzzy join)
  
# Noticeably, regions 13 and 4 had the largest data quality 
# problem here.
  
remove_name <- function(school, region){
  tem <- subset(vase, !(School == school & Region == region))
  return(tem)
}

vase <- remove_name("Reagan H S", 13)
vase <- remove_name("Lanier H S", 13)
vase <- remove_name("Seguin H S", 13)
vase <- remove_name("Robert E Lee H S", 7)
vase <- remove_name("Robert E. Lee H S", 7)
vase <- remove_name("James Nikki Rowe H S", 1)
vase <- remove_name("Dekaney", 4)
vase <- remove_name("Connally H S", 13)
vase <- remove_name("Lee H S", 18)
vase <- remove_name("Canyon Lake H S", 13)
vase <- remove_name("St. Pius X", 4)
vase <- remove_name("Falcon ECHS", 4)
vase <- remove_name("Falcon ECHS", 18)
vase <- remove_name("Gsisd", 7)
vase <- remove_name("Private", 3)
vase <- remove_name("Lisd", 5)
vase <- remove_name("Kolarik", 4)
vase <- remove_name("Duchesne Academy", 4)
vase <- remove_name("Episcopal H S", 4)
vase <- remove_name("Canyon H S", 13)
vase <- remove_name("Davenport H S", 13)
vase <- remove_name("H C T C", 11)
vase <- remove_name("Jefferson H S", 14)
vase <- remove_name("Keystone School", 20)
vase <- remove_name("Northland", 4)


# Multiple schools with same name, but missing further identifiers (i.e. suffix or district)
vase <- remove_name("Kipp", 4)
vase <- remove_name("Sterling", 4)
vase <- remove_name("Steele H S", 20)
vase <- remove_name("Steele", 20)

```

**Fuzzy Join**: Due to large size of `vase`, the dataset needs to be collapsed for fuzzy join to run quickly without overloading memory. However, quite a few schools have multiple, inconsistent formats in `vase`, such as not incorporating H S or other important school type identifiers. Further, name lengths are dramatically different between `vase` and `tea22`, which greatly affects matching methods.  

So a smaller dataframe will be created and joined with `tea22`. After initial fuzzyjoin, matches with high distances will be manually reviewed and edited. Non-unique school names will be cross referenced in `tea22` to find extra identifiers. Then `vase` will be joined with this dataframe.

```{r}

# get values using school and region
keys <- distinct(vase, School, Region)

# fuzzy join and take only the best match (if one exists)
dict <- fuzzyjoin::stringdist_left_join(keys, tea22, 
                                        by = c("School", "Region"),
                                        ignore_case = TRUE, 
                                        method = "lcs",
                                        max_dist = 50,
                                        distance_col = "dist")
dict <- dict %>%
  filter(Region.dist == 0) %>%
  group_by(School.x, Region.x) %>%
  slice_min(order_by = School.dist, n = 1) %>%
  arrange(desc(School.dist))

# Same school names in same region, but different districts.

# create function that will correct data entries for duplicate entries that have unique extra identifiers
rplc <- function(left, right, isd, rgn){
  temp <- dict

  for (i in 3:6){
    temp[temp$School.x == left & temp$Region.x == rgn, 3:6] <- tea22[tea22$School == right & tea22$District == isd]
  }
  
  # set distance to 0 since it does have a perfect match
  temp[temp$School.x == left & temp$Region.x == rgn, 8] <- 0
  
  return(temp)
}

dict <- rplc("Alief Taylor", "TAYLOR H S", "ALIEF ISD", 4)
dict <- rplc("Alief Taylor H S", "TAYLOR H S", "ALIEF ISD", 4)
dict <- rplc("Aldine Nimitz", "NIMITZ H S", "ALDINE ISD", 4)
dict <- rplc("James E. Taylor", "TAYLOR H S", "KATY ISD", 4)

dict <- dict %>%
  filter(School.dist <= 10) %>%
  drop_na(School.x)
```

**Create Dataset** The updated `vase` dataset will be created using `dict`. Then the `.x` columns will be removed so only standardized values remaining.

```{r}

```